A universal of human social cognition: Children from 17 communities process gaze in similar ways

### 2025-02-26 paper/paper.tex
add agent info to the data

\abstract{%
Theoretical accounts assume that key features of human social cognition are universal. Here we focus on gaze following, the bedrock of social interactions and coordinated activities, to test this claim. In a comprehensive cross-cultural study spanning five continents and 17 distinct cultural communities, we examined the development of gaze following in early childhood. We identified key processing signatures through a computational model that assumes that participants follow an individual's gaze by estimating a vector emanating from the eye center through the pupil. We found these signatures in all communities, suggesting that children worldwide processed gaze in highly similar ways. Absolute differences between groups were accounted for by a cross-culturally consistent relationship between children's exposure to touchscreens and their performance in the task. These results provide strong evidence for a universal process underlying a foundational socio-cognitive ability in humans that can be reliably inferred even in the presence of cultural variation in overt behavior.
}

@@

\begin{verbatim}
## # A tibble: 2 x 3
##   side  left_bias right_bias
##   <chr>     <dbl>      <dbl>
## 1 left      0.170      0.830
## 2 right     0.754      0.246
\end{verbatim}

@@

Human socio-cognitive skills enable unique forms of communication and cooperation that provide a bedrock for cumulative culture and the formation of complex societies (Henrich, 2016; Heyes, 2018; Laland \& Seed, 2021; Legare, 2019; Tomasello, 2020; Tomasello \& Rakoczy, 2003; Wellman, 2014). The eyes are the proverbial ``window to the mind'' and eye gaze is essential for many social reasoning processes (Doherty, 2006; Emery, 2000; Frischen et al., 2007; Shepherd, 2010). Others' eye gaze is used to infer their focus of visual attention, which is a critical aspect of coordinated activities, including communication and cooperation (Langton et al., 2000; Richardson \& Dale, 2005; Rossano, 2012; Scaife \& Bruner, 1975; Sebanz et al., 2006; Tomasello et al., 2007). During ontogeny, gaze following is an important aspect of many critical learning objectives that enable children to become functioning members of the society they grow up in, including language, social learning and joint action (Brooks \& Meltzoff, 2005; Brownell, 2011; Carpenter et al., 1998; Moore, 2008; Mundy \& Newell, 2007; Stephenson et al., 2021). Because of the central role gaze following plays during human ontogeny, it has been widely argued that gaze following has been a target of natural selection (Clark et al., 2023; Emery, 2000; Kano, 2023; Tomasello et al., 2007). This implies that the process by which humans use gaze direction to infer the focus of attention is universal. In this paper, we report a comprehensive cross-cultural study on the ontogeny of gaze following in which we shed light on the universal aspects of gaze following as well as sources of variation and their origins.

\hypertarget{ontogeny-of-gaze-following}{%
\subsection{Ontogeny of gaze following}\label{ontogeny-of-gaze-following}}

The ability to follow gaze emerges early in development (Byers-Heinlein et al., 2021; Del Bianco et al., 2019; Gredebäck et al., 2010; Tang et al., 2023). The earliest signs of gaze following have been found in infants as young as four months (Astor et al., 2021; D'Entremont et al., 1997). Throughout the first two years of life, children refine their abilities: they interpret gaze in mentalistic terms (Butterworth \& Jarrett, 1991; Deák et al., 2000), for example, they follow gaze to locations outside their own visual field by moving around barriers (Moll \& Tomasello, 2004). Initially, children rely more on head direction than actual gaze direction (Michel et al., 2021). In fact, when head and gaze direction diverge, children fail to accurately locate the agent's focus of attention up until 19 months of age (Lempers, 1979).

From an evolutionary perspective, while many species are able to follow gaze based on head directions, uniquely human forms of joint action and communication require a more precise localization of other's attention and thus critically rely on gaze direction inferred from eye movements (Emery, 2000; Hessels, 2020). In a recent study, Prein, Maurits, et al. (2024) studied the development of gaze following based on eye movements across from three years up until old age. They found particularly steep developmental improvements in the preschool years resulting in a relatively stable level of accuracy from ten years onward and a slight decrease starting around age 40.

The studies reported thus far, all relied on data collected in western affluent settings. Such settings represent only a minority of the worlds population and are thus insufficient to make claims about universal aspects of human cognition (Amir \& McAuliffe, 2020; Nielsen et al., 2017; Norenzayan \& Heine, 2005). Three studies with infants and children from traditionally underrepresented parts of the world (Bhutan, India, Peru, Vanuatu) find that children start gaze following (including head direction) at similar ages, the rates of gaze following, however, differed between cultural settings (Astor et al., 2022; Callaghan et al., 2011; Hernik \& Broesch, 2019).

Rates and accuracy of gaze following do not just differ between cultural settings; there is also substantial variation within settings. In fact, the pivotal role of gaze following in many uniquely human activities has been studied by relating individual differences in gaze following to other phenomena -- both cross-sectionally and longitudinally (Brooks \& Meltzoff, 2015; Carpenter et al., 1998). For example, gaze following at 10 months predicts language scores at 18 months of age (Brooks \& Meltzoff, 2005; see also Macdonald \& Tatler, 2013). Furthermore, difficulties with gaze following have been linked to developmental disorders, including Autism (Itier \& Batty, 2009; Thorup et al., 2016, 2018) and -- at least in some cultural contexts -- to maternal postpartum depression {[}astor2022maternal{]}. Individual differences are also key to explaining the -- mainly social-interactional -- driving forces behind the development of gaze following. For example, it has been found that early attachment quality or the use of gaze in communicative interactions predict later rates of gaze following (Astor et al., 2020; Movellan \& Watson, 2002; Senju et al., 2015).

\hypertarget{cognitive-universals-and-sources-of-variation}{%
\subsection{Cognitive universals and sources of variation}\label{cognitive-universals-and-sources-of-variation}}

The existence of variation in gaze following both within and between cultural settings raises the question of how to square these findings with the suggestion that gaze following is a fundamental building block of human social cognition and interaction and that eye movements are processed the same way in humans all over the world. Looking at other aspects of social cognition, one could easily make the argument that variation is the norm rather than the exception (Dixson et al., 2018; Mayer \& Träuble, 2013; see e.g., Miller et al., 2018; Taumoepeau et al., 2019; Wellman, 2014). As a first step, answering this question requires data from many different cultural settings. In a second step, however, we need to a way of detecting universal processes in such data. The traditional approach is to compare some sort of aggregate measure (mean level of performance, average age of onset) across cultural settings. Absolute differences in mean performance across communities are interpreted as a signal of different underlying cognitive processes (Blake et al., 2015; see e.g., House et al., 2020; Kanngiesser et al., 2022; Van Leeuwen et al., 2018). No differences seemingly support the existence of a psychological universal. Such an approach, however, neglects the existence of within-cultural variation altogether (see also Gurven, 2018).

In the present study, we want to take a different approach for which we assume that universal processes and variation can co-exist (Greenfield et al., 2003; Jensen, 2012; Kline et al., 2018). Instead of starting with the outcome (performance in the task), we start with the process that generates the outcome. By defining this process, we make a proposal for the universal aspect of the process. At the same time, we define variable aspects in the process that generate individual differences. This allows us to define signatures that the process leaves behind that can be detected independent of absolute levels of performance.

For gaze following, we can use the computational model proposed by Prein, Maurits, et al. (2024) to derive such predictions. They formalized the widely-held view that gaze following involves estimating a vector emanating from the eye center through the pupil (Butterworth \& Jarrett, 1991; Symons et al., 2004; Todorović, 2006; Yaniv \& Shatz, 1990). The key innovation of the model is that it explains how individuals may use the same process but still differ in their measured abilities. The process always involves estimating a vector but also involves a degree of uncertainty because the eye center is not directly observable. Individuals are assumed to differ in their level of uncertainty with which they estimate the vector which causes differences in their observable behavior. Importantly, the assumed process leaves a key signature in the data that is observable independent of the absolute level of performance. In the present study, we therefore focus on this signature instead of absolute levels of performance when evaluating the claim whether there is evidence for a universal cognitive process underlying gaze following.

@@

The present study had three goals. First, to collect a comprehensive data set and study the ontogeny of gaze following beyond infancy across cultures. To make this possible, we used a semi-standardized task that required minimal assistance from an experimenter and no behavioral coding. The task is an animated picture book presented on a tablet screen. Children watched a balloon disappear behind a hedge. An agent followed the trajectory of the balloon with their eyes (Figure \ref{fig:fig1}B). The key dependent variable was (im)precision, that is, the deviation between where the agent looked (where the balloon was) and the child's response. The task's flexible implementation as a browser-based web-app allowed us to quickly tailor its visual and audio content to each cultural setting (visuals and audio). The task has been psychometrically evaluated and has shown to yield reliable individual-level measurements across communities and across ages (Prein, Kalinke, et al., 2024; Prein, Bednarski, et al., 2024).

We collected data in 17 different communities across 14 countries and five continents. Communities covered a broad spectrum of geographical locations, social and political systems, languages, and subsistence styles. This diversity allowed us to overcome a common pitfall of cross-cultural studies that compare urban communities from the Global North to rural communities from the Global South (Barrett, 2020). We aimed for large sample sizes within each community to contrast within- and between cultural variation. Our expectation regarding the first goal was to see substantial variation across cultures but even more variation between individuals. In all communities, we expected performance to increase with age.

The second goals was to look for signatures in the data of the universal gaze following process predicted by the model of Prein, Bednarski, et al. (2024). In the task, the hidden object is located on a horizontal plane at the lower end of the screen. The agent is located in the upper center of the screen (see Figure \ref{fig:fig1}B). The model predicts trials in which the object is hidden further away from the center to be more difficult resulting in higher imprecision. The signature is thus a u-shaped relation between object location and imprecision (Figure \ref{fig:fig12}).

Finally, we sought to explain individual differences in gaze following precision by linking them to methodological aspects of the study as well as aggregate measures of children's everyday social experience. Experience with tablets and touch screens co-varied with community and we expected children more familiar with this medium to perform better. Previous work suggested that gaze following is refined in social interaction (Movellan \& Watson, 2002; Senju et al., 2015). To approximate social interaction, we asked parents to fill out a questionnaire about household size and composition. We acknowledge that this measure approximates opportunities for social interaction in a rather crude way but we nevertheless expected children living in larger households and with more siblings (relative to their community) to be more accurate when following gaze.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{../figures/fig1_2}

\caption{(A) Data collection sites. Points show the approximate geographical location of the data collection sites, coloring shows the sample sizes. (B) Screenshots from the task. The upper scene depicts the start, and the lower scene depicts the choice phase in a test trial. Participants had to use the gaze of the agent to locate the balloon and touch the location on the hedge where they thought the balloon was. Agents, audio recordings and backgrounds were adapted to each community. (C) Drawings used as agents across communities.}\label{fig:fig1}
\end{figure}

@@

We used an animated picture book tablet task in which participants had to locate a hidden object based on observing an agent's gaze. Children watched a balloon disappear behind a hedge. An agent followed the trajectory of the balloon with their eyes (Fig. \ref{fig:fig1}B). The key dependent variable was the (im)precision with which children located the agent's focus of attention, that is, the deviation between where the agent looked (where the balloon was) and the child's response. We adapted visuals and audio instructions specifically for each of the 17 communities. Previous work demonstrated excellent individual-level measurement properties for this task in a German sample (Prein, Kalinke, et al., 2024).

@@

@@

The study design, the sampling strategy and the general analytic strategy were preregistered prior to data collection (\url{https://osf.io/tdsvc}). The final sample size was not preregistered because we did not know how many communities would participate when the study began. Instead, we stated the age range we intended to study in each community (3.0 to 5.9 years of age) along and that we planned to test 20 children per year bin. We achieved this goal for most ages in most communities, but not for all (see Supplementary Table 1). The analysis reported here deviated from the preregistration in the following ways: In the pre-registration, in the regression models, we did not include random slopes for target centrality (i.e.~the distance of the location where the balloon landed from the center of the screen), neither within subject nor within cultural setting. Instead, we included random slopes for trial within subject. We decided to change this and remove random slopes for trial but include them for target centrality because trial effects in the sense of learning across trials were unlikely because participants did not receive differential feedback. Furthermore, multiple studies with the same task since the registration found no trial effects (Prein, Kalinke, et al., 2024; Prein, Maurits, et al., 2024). On the other hand, we included random slopes for target centrality to be able to look at cross-cultural variation. The cognitive models were not mentioned in the pre-registration because it is a more recent development (Prein, Maurits, et al., 2024).

@@

All study materials (\url{https://ccp-odc.eva.mpg.de/tango-cc/}), primary data and analysis scripts are publicly available (\url{https://github.com/ccp-eva/gafo-cc-analysis/}).

@@

A total of 1377 children between 2.38 and 10.95 years of age provided data for the study. Children lived in 17 different communities, located in 14 different countries across five continents. Table \ref{tab:tab1} gives the sample size per community together with basic demographic information and age. For some children, the exact birthday was unknown. In such cases, we set the birthday to the 30th of June of the year that would make them fall into the reported age category. We provide a detailed description of the sample characteristics, the study site and recruitment strategy for each community in the Supplementary Material.

Data from children was only included in the study when they contributed at least four valid test trials. We also excluded the data from children with a diagnosed developmental disorder. In sum, in addition to the sample size reported above, 74 additional children participated in the study but did not contribute data. The main reasons for exclusion were: contribution of less than four valid test trials, technical failures, and missing or implausible demographic information (e.g., when the number of children living in the household was reported to be larger than the household itself or when the number of children reported to live in the household equaled the number of children younger than the child being tested). We did not exclude any participants for performance reasons. A detailed description of each data collection site and the way children were recruited can be found in the Supplementary Material.

@@

The task was implemented as a browser-based interactive picture book using \texttt{HTML}, \texttt{CSS}, and \texttt{JavaScript}. Participants saw animated agents on a touch screen device, listened to pre-recorded audio instructions and responded by touching the screen. In all communities, a research assistant, fluent in the local language(s), guided the child through the task. That is, the research assistant guided the child through the introduction and advanced the study from trial to trial.

Figure \ref{fig:fig1}B shows a screenshot from the task. The task was introduced verbally by the assistant as the balloon game in which the participant would play with other children to find a balloon. On each trial, participants saw an agent located in a window in the center of the screen. A balloon fell down from its starting position just below the agent. The agent's gaze followed the trajectory of the balloon. That is, the pupils and the iris were programmed to align with the center of the balloon. Once the balloon had landed on the ground, the agent was instructed to locate it, that is, to touch the location on the screen where they thought the balloon was. On each trial, we recorded the exact x-coordinate of the participant's touch.

There were two types of training trials. In training 1 trials, the balloon fell down and landed in plain sight. Participants simply had to touch the visible balloon. In training 2 trials, the trajectory of the balloon was visible but it landed behind a small barrier (a hedge -- see Figure \ref{fig:fig1}B). Thus, participants needed to touch the hedge where they saw the balloon land. Next came test trials. Here, the barrier moved up and covered the balloon's trajectory. That is, participants only saw the agent's eyes move, but not the balloon. They had to infer the location of the balloon based on the agent's gaze direction. During training 1, training 2 and the first test trial, children heard voice-overs commenting what happened on the screen. Critically, the agent was described as wanting to help the child and always looking at the balloon.

Children completed one training 1, two training 2 trials and 16 test trials. We excluded the first test trial from the analysis because of the voice-over. Thus, 15 test trials were used in the analysis below. Each child saw eight different agents (four male, four female). The agent changed from trial to trial, with alternating genders. A coin toss before the first trial decided whether the first agent was male or female. The order in which agents were shown was randomized with the constraint that all agents had to be shown once until an agent was shown again. The color of the balloon also changed from trial to trial in a random order, also with the constraint that all colors appeared once before any one was repeated.

The location (x-coordinate) where the balloon landed was determined in the following way: The screen was divided in ten equally sized bins. On each trial, one of the bins was randomly selected and the exact x-coordinate was randomly chosen within that bin. Constraints were that the balloon landed in each bin once in the first ten trials and, for the remaining six test trials, it landed in a different om each trial. Thus, each bin appeared a no more than twice.

All children were tested with a touchscreen device with a size between 11 and 13 inch equipped with a webcam. The data was either stored locally or sent to a server. In addition to the behavioral data, we stored the webcam recording of the session for verification purposes. Community-specific adaptations were made by changing the visuals and the audio instructions (see Supplementary Material for details).

@@

The first analysis was focused on cross-cultural variation. Fixed effects in the model were age and target centrality (distance of the landing position from the center in pixel/SVG units). The latter term accounts for trial difficulty. Furthermore, we included participant as a random effect, with a random slope for target centrality. To assess cross-cultural variation, we compared three models. A null model without cultural setting as a predictor (\texttt{brms} notation: \texttt{imprecision\ \textasciitilde{}\ age\ +\ target\_centrality\ +\ (target\_centrality\ \textbar{}\ participant)}), a model with cultural setting as a random intercept (\texttt{imprecision\ \textasciitilde{}\ age\ +\ target\_centrality\ +\ (target\_centrality\ \textbar{}\ participant)\ +\ (target\_centrality\ \textbar{}\ community)}) and a model with cultural setting as a random intercept and an added random slope for age (\texttt{imprecision\ \textasciitilde{}\ age\ +\ target\_centrality\ +\ (target\_centrality\ \textbar{}\ participant)\ +\ (age\ +\ target\_centrality\ \textbar{}\ community)}). Thus, the second model assumes that there is variation across cultures in average levels of precision and the third model assumes that there are additional cultural differences in the effect of age.

As stated in the pre-registration, comparing these models could be problematic. Participants are fully nested within cultural setting. If there was an effect of cultural setting, we would expect participant random intercepts to cluster by cultural setting. This clustering would appear whether or not cultural setting would be included in the model as a random effect or not -- the only difference would be if the participant random intercepts were estimated as a deviation from a grand intercept or a culture-specific one. Standard metrics such as \texttt{WAIC} or \texttt{LOO} would penalize the model with additional intercept for cultural setting for having additional parameters that do not help to improve predictive accuracy.

To get around this problem, we used a cross-validation procedure (see e.g., 6). For each cultural setting, we randomly sampled a data set that was 5/6 the size of the full data set (training data). Then, we fit the model to this training data and used the estimated model parameters to predict the remaining 1/6 of the data (testing data). We then compared the model predictions from the different models by computing the mean difference between the true and predicted imprecision, over all trials in the testing data set. This approach gets around the problem mentioned above because the model predicts a new data set for which the individual random intercepts are unknown. Clustering by culture could therefore only be predicted by a model that included culture as a predictor. We repeated the cross-validation procedure 100 times and counted which model performed best most often.

@@

\begin{figure}

>>>>

\abstract{%
Theoretical accounts typically assume, but rarely test, that key features of human socio-cognitive development are universal. This paper reports a comprehensive cross-cultural study (17 communities, five continents, N = 1377, 709 female, mean = 5.50 years ) on gaze following in early childhood. To test for universality, cognitive processing signatures were derived from a computational model treating gaze following as social vector estimation. Results showed substantial variation between communities and individuals. Importantly, the processing signature was found in all communities. Individual differences in performance were related to individual-level measures of familiarity with the data-collection device but not opportunities for social interaction. These results provide strong evidence for gaze following as a universal socio-cognitive process despite cultural variation in overt behavior.
}

@@

@@

Human socio-cognitive skills enable unique forms of communication and cooperation that provide a bedrock for cumulative culture and the formation of complex societies (Henrich, 2016; Heyes, 2018; Laland \& Seed, 2021; Legare, 2019; Tomasello \& Rakoczy, 2003). Eye gaze is essential for many social reasoning processes, making the eyes the proverbial `'window to the mind'' (Doherty, 2006; Emery, 2000; Frischen et al., 2007; Shepherd, 2010). Others' eye gaze is used to infer their focus of visual attention, which is a critical aspect of coordinated activities, including communication and cooperation (Langton et al., 2000; Richardson \& Dale, 2005; Rossano, 2012; Scaife \& Bruner, 1975; Sebanz et al., 2006; Tomasello et al., 2007). During ontogeny, gaze following is an important aspect of many critical learning objectives that enable children to become functioning members of the society they grow up in, including language, social learning and cooperation (Brooks \& Meltzoff, 2005; Brownell, 2011; Carpenter et al., 1998; Moore, 2008; Mundy \& Newell, 2007; Stephenson et al., 2021). Because of the central role gaze following plays during human ontogeny, it has been widely argued that gaze following has been a target of natural selection (Clark et al., 2023; Emery, 2000; Kano, 2023; Tomasello et al., 2007). This implies that the process by which humans use gaze direction to infer the focus of attention is universal. In this paper, we report a comprehensive cross-cultural study on the ontogeny of gaze following in which we shed light on the universal aspects of gaze following as well as sources of variation and their origins.

The ability to follow gaze emerges early in development (Byers-Heinlein et al., 2021; Del Bianco et al., 2019; Gredebäck et al., 2010; Tang et al., 2023). The earliest signs of gaze following have been found in infants as young as four months (Astor et al., 2021; D'Entremont et al., 1997). Initially, children rely more on head direction than actual gaze direction (Michel et al., 2021). When head and gaze direction diverge, children often fail to accurately locate the agent's focus of attention up until 19 months of age (Lempers, 1979). Throughout the first two years of life, children refine their abilities: they begin to interpret gaze in mentalistic terms (Butterworth \& Jarrett, 1991; Deák et al., 2000), for example, they follow gaze to locations outside their own visual field by moving around barriers (Moll \& Tomasello, 2004).

From an evolutionary perspective, while many species are able to follow gaze based on head directions, uniquely human forms of joint action and communication require a more precise localization of other's attention and thus critically rely on gaze direction inferred from eye movements (Emery, 2000; Hessels, 2020). In a recent study, Prein, Maurits, et al. (2024) studied the development of gaze following based on eye movements from three years up until old age. They found particularly steep developmental improvements in the preschool years resulting in a relatively stable level of accuracy from ten years onward and a slight decrease starting around age 40.

The studies reported thus far, relied on data collected in western affluent communitites. Such communitites represent only a minority of the worlds population and are thus insufficient to make claims about universal aspects of human cognition and its development (Amir \& McAuliffe, 2020; Barrett, 2020; Nielsen et al., 2017; Norenzayan \& Heine, 2005). Three studies with infants and children from traditionally underrepresented regions (Bhutan, India, Peru, Vanuatu) find that even though children start gaze following (gaze and head direction combined) at similar ages, the rates of gaze following differ between cultural communitites (Astor et al., 2022; Callaghan et al., 2011; Hernik \& Broesch, 2019).

Rates and accuracy of gaze following do not just differ between cultural communitites; there is also substantial variation within communitites. In fact, the pivotal role of gaze following in many uniquely human activities has been studied by relating individual differences in gaze following to other phenomena -- both cross-sectionally and longitudinally (Brooks \& Meltzoff, 2015; Carpenter et al., 1998). For example, gaze following at 10 months predicts language scores at 18 months of age (Brooks \& Meltzoff, 2005; Macdonald \& Tatler, 2013). Furthermore, difficulties with gaze following have been linked to developmental disorders, including Autism (Itier \& Batty, 2009; Thorup et al., 2016) and -- at least in some cultural contexts -- to maternal postpartum depression (Astor et al., 2022). Individual differences are also key to explaining the driving forces behind the development of gaze following. For example, it has been found that early attachment quality or the use of gaze in communicative interactions predict later rates of gaze following (Astor et al., 2020; Movellan \& Watson, 2002; Senju et al., 2015).

The existence of variation in gaze following both within and between cultural communities raises the question of how to square these findings with the suggestion that gaze following is a fundamental building block of human social cognition and interaction and that humans all over the world process eye movements in the same way. Looking at other areas of social cognition, one could easily make the argument that variation is the norm rather than the exception (see e.g., Dixson et al., 2018; Kaminski et al., 2024; Mayer \& Träuble, 2013; Miller et al., 2018; Stengelin et al., 2020, 2024; Taumoepeau et al., 2019). As a first step, answering this question requires data from many different cultural communitites. In a second step, however, we need a way of detecting potential universal processing signatures in such data. Aggregate measures (e.g., mean level of performance, average age of onset) are often compared across cultural communities and absolute differences between communities are interpreted as a signal of different underlying cognitive processes while no differences are taken to support the existence of a psychological universal (Blake et al., 2015; House et al., 2020; Kanngiesser et al., 2022; Van Leeuwen et al., 2018). Such an approach -- while helpful in broadly mapping cross-cultural variation and similarities -- cannot account for potential within-cultural variation (see also Gurven, 2018).

In the present study, we take a different approach that assumes the potential for universal processes and variation to co-exist (Greenfield et al., 2003; Jensen, 2012; Kline et al., 2018). Instead of starting with the outcome (performance in the task), we begin with the process that generates the outcome and propose universal signatures. At the same time, we define how individuals may differ from one another when using the process. This allows us to define signatures that the process leaves behind in the data that can be detected independent of individual-level or cross-cultural variation in absolute performance.

The computational model proposed by Prein, Maurits, et al. (2024) can be used to derive such predictions for gaze following. They formalized the widely-held view that gaze following involves estimating a line-of-sight vector emanating from the eye center through the pupil (Butterworth \& Jarrett, 1991; Todorović, 2006; Yaniv \& Shatz, 1990). The key innovation of the model is that it explains how the same underlying process can produce different behavioral outcomes. The process always involves estimating a vector with a variable degree of uncertainty because the eye center is not directly observable. Individuals are assumed to differ in their level of uncertainty when estimating the vector which causes differences in their observable behavior. Importantly, this general vector estimation process emerges as a key signature in the data that is detectable independent of the absolute level of performance. In the present study, we therefore focus on this signature instead of absolute levels of performance when evaluating whether there is evidence for a universal cognitive process underlying gaze following.

@@

The present study had three goals. First, to collect a comprehensive data set and study the ontogeny of gaze following beyond infancy across cultures. To make this possible, we used a semi-standardized task that required minimal assistance from an experimenter and no behavioral coding {[}TANGO-CC; Prein, Bednarski, et al. (2024){]}. The task is an animated picture book presented on a tablet screen. Children watched a balloon disappear behind a hedge. An agent followed the trajectory of the balloon with their eyes (Figure \ref{fig:fig1}B). The key dependent variable was (im)precision, that is, the deviation between where the agent looked (where the balloon was) and the child's response. The task's flexible implementation as a browser-based web-app allowed us to quickly tailor its visual and audio content to each cultural community (visuals and audio). Adaptations were made by researchers or research assistants from the respective community. The task has been psychometrically evaluated and has shown to yield reliable individual-level measurements across communities and ages (Prein, Kalinke, et al., 2024; Prein, Bednarski, et al., 2024). Furthermore, it has been validated in a German sample predicting language abilities six months later and correlating with conventional real-life perspective taking tasks (Prein, Kalinke, et al., 2024, 2024; Prein, Maurits, et al., 2024).

We collected data in 17 different communities across 14 countries and five continents (see Figure \ref{fig:fig1}A). Communities covered a broad spectrum of geographical locations, social and political systems, languages, and subsistence styles. This diversity allowed us to overcome a common pitfall of cross-cultural studies that compare urban communities from the Global North to rural communities from the Global South (Barrett, 2020). We aimed for large sample sizes within each community (n = 20 per year) to contrast within- and between cultural variation. Our expectation regarding the first goal was to see substantial variation across cultures but even more variation between individuals. In all communities, we expected performance to improve with age.

The second goal was to look for signatures in the data of the universal gaze following process specified by the model of Prein, Maurits, et al. (2024). In the task, the hidden object lands on a horizontal plane at the lower end of the screen. The agent is located in the upper center of the screen (see Figure \ref{fig:fig1}B). The model predicts trials in which the object is hidden further away from the center to be more difficult, resulting in higher imprecision. The signature is thus a u-shaped relation between object location and imprecision (Figure \ref{fig:fig12}).

Finally, we sought to explain individual differences in gaze following precision by linking them to methodological aspects of the study as well as aggregate measures of children's everyday social experience. Experience with tablets and touch screens co-varied with community and we expected children more familiar with these devices to perform better on tablet-based tasks. Previous work suggested that gaze following is refined in social interaction (Movellan \& Watson, 2002; Senju et al., 2015). To approximate social interaction, we asked parents to fill out a questionnaire about household size and composition. We acknowledge that this measure approximates opportunities for social interaction in a rather coarse way but we nevertheless expected children living in larger households and with more siblings (relative to their community) to be more accurate when following gaze.

\begin{figure}
\includegraphics[width=1\linewidth]{../figures/fig1_2} \caption{(A) Data collection sites. Points show the approximate geographical location of the data collection sites, coloring shows the sample sizes. (B) Screenshots from the task. The lower scene depicts the choice phase in a test trial. Participants had to use the gaze of the agent to locate the balloon and touch the location on the hedge where they thought the balloon was. Agents, audio recordings and backgrounds were adapted to each community. (C) Drawings used as agents across communities. Adaptations were made by researchers or research assistants from the respective community.}\label{fig:fig1}
\end{figure}

@@

@@

The study obtained ethical clearance from the MPG Ethics Commission in Munich, Germany, falling under an umbrella ethics application (Appl. No.~2021\_45).

@@

The study design, the sampling strategy and the general analytic strategy were preregistered prior to data collection (\url{https://osf.io/tdsvc}). The final sample size was not preregistered because we did not know how many communities would participate when the study began. Instead, we stated the age range we intended to study in each community (3.0 to 5.9 years of age) along and that we planned to test 20 children per year bin. We achieved this goal for most ages in most communities, but not for all (see Supplementary Table 1). For some communities, we also included older and younger children. The analysis reported here deviated from the preregistration in the following ways: In the preregistration, the regression models did not include random slopes for target centrality (i.e., the distance between the location where the balloon landed and the center of the screen), neither within subject nor within cultural setting. Instead, the preregistration included random slopes for trial within subject. We decided to remove random slopes for trial but include them for target centrality because trial effects in the sense of learning across trials were unlikely in the absence of differential feedback. Furthermore, multiple studies with the same task since the registration found no trial effects (Prein, Kalinke, et al., 2024; Prein, Maurits, et al., 2024). On the other hand, we included random slopes for target centrality to be able to look at cross-cultural variation. We preregistered to run the cross-validation 1000 times but decided to scale down to 100 times because the procedure was computationally very intensive and the results very clear. The cognitive models were not mentioned in the preregistration because they were developed more recently (Prein, Maurits, et al., 2024).

@@

All study materials (\url{https://ccp-odc.eva.mpg.de/tango-cc/}), primary data, and analysis scripts are publicly available (\url{https://github.com/ccp-eva/gafo-cc-analysis/}).

@@

A total of 1377 children between 2.38 and 10.95 years of age provided data for the study. Children lived in 17 different communities, located in 14 different countries across five continents. Table \ref{tab:tab1} gives the sample size per community together with basic demographic information and age. For some children, the exact birthday was unknown. In such cases, we set the birthday to the 30th of June of the year that would make them fall into the reported age category. We provide a detailed description of the sample characteristics, the age distributions, the study sites and recruitment strategies for each community in the Supplementary Material. Samples were convenience samples in all communities.

Data from children was only included in the study when they contributed at least four valid test trials. We also excluded the data from children when a developmental disorder was reported. In addition to the sample size reported above, 74 children participated in the study but did not contribute data. The main reasons for exclusion were: contribution of less than four valid test trials, technical failures, and missing or implausible demographic information (e.g., when the number of children living in the household was reported to be larger than the household itself or when the number of children reported to live in the household equaled the number of children younger than the child being tested). We did not exclude any participants for performance reasons.

@@

The task was implemented as a browser-based interactive picture book using \texttt{HTML}, \texttt{CSS}, and \texttt{JavaScript}. Participants saw animated agents on a touch screen device, listened to pre-recorded audio instructions and responded by touching the screen. In all communities, a research assistant, fluent in the local language(s), guided the child through the introduction and advanced the study from trial to trial.

Figure \ref{fig:fig1}B shows a screenshot from the task. The task was introduced verbally by the assistant as the balloon game in which the participant would play with other children to find a balloon. On each trial, participants saw an agent located in a window in the center of the screen. A balloon fell down from its starting position just below the agent. The agent's gaze followed the trajectory of the balloon. That is, the pupils and the iris were programmed to align with the center of the balloon. Once the balloon had landed on the ground, the child was instructed to locate it, that is, to touch the location on the screen where they thought the balloon was. On each trial, we recorded the exact x-coordinate of the participant's touch.

There were two types of training trials. In no-hedge training trials, the balloon fell down and landed in plain sight. Participants simply had to touch the visible balloon. In hedge training trials, the trajectory of the balloon was visible but it landed behind a small barrier (a hedge -- see Figure \ref{fig:fig1}B). Thus, participants needed to touch the hedge where they saw the balloon land. Next came test trials. Here, the barrier moved up and covered the balloon's trajectory. That is, participants only saw the agent's eyes move, but not the balloon. They had to infer the location of the balloon based on the agent's gaze direction. During training and the first test trial, children heard voice-overs commenting what happened on the screen. Critically, the agent was described as wanting to help the child and always looking at the balloon. These instructions were added to clarify the purpose of the task, establish a clear common ground, and minimize learning effects over trials.

Children completed one no-hedge training, two hedge training trials and 16 test trials. We excluded the first test trial from the analysis because of the voice-over. Thus, 15 test trials were used in the analysis below. Each child saw eight different agents (four male, four female; selected by local researchers or research assistants). The agent changed from trial to trial, with alternating genders. A coin toss before the first trial decided whether the first agent was male or female. The order in which agents were shown was randomized with the constraint that all agents had to be shown once until an agent was shown again. The color of the balloon also changed from trial to trial in a random order, also with the constraint that all colors appeared once before any one was repeated.

The location (x-coordinate) where the balloon landed was determined in the following way: The screen was divided in ten equally sized bins. On each trial, one of the bins was randomly selected and the exact x-coordinate was randomly chosen within that bin. Constraints were that the balloon landed in each bin once in the first ten trials and, for the remaining six test trials, it landed in a different bin on each trial. Thus, each bin appeared no more than twice.

All children were tested with a touchscreen device with a size between 11 and 13 inch equipped with a webcam. The data was stored locally. In addition to the behavioral data, we stored the webcam recording of the session for verification purposes. Community-specific adaptations were made by changing the visuals and the audio instructions (see Supplementary Material for details).

@@

The first analysis was focused on cross-cultural variation. Fixed effects in the model were age and target centrality (distance of the balloon's landing position from the center in pixel/SVG units). The latter term accounts for trial difficulty (see below). Furthermore, we included participant as a random effect, with a random slope for target centrality. To assess cross-cultural variation, we compared three models: a null model without cultural community as a predictor (\texttt{brms} notation: \texttt{imprecision\ \textasciitilde{}\ age\ +\ target\_centrality\ +\ (target\_centrality\ \textbar{}\ participant)}), a model with cultural community as a random intercept (\texttt{imprecision\ \textasciitilde{}\ age\ +\ target\_centrality\ +\ (target\_centrality\ \textbar{}\ participant)\ +\ (target\_centrality\ \textbar{}\ community)}), and a model with cultural community as a random intercept and an added random slope for age (\texttt{imprecision\ \textasciitilde{}\ age\ +\ target\_centrality\ +\ (target\_centrality\ \textbar{}\ participant)\ +\ (age\ +\ target\_centrality\ \textbar{}\ community)}). Thus, the second model assumes that there is variation across cultures in average levels of precision and the third model assumes that there are additional cultural differences in the effect of age.

As stated in the preregistration , comparing these models could be problematic. Participants are fully nested within a cultural setting. If there was an effect of cultural setting, we would expect participant random intercepts to cluster by cultural setting. This clustering would appear whether or not cultural community would be included in the model as a random effect or not -- the only difference would be if the participant random intercepts were estimated as a deviation from a grand intercept or a culture-specific one. Standard metrics such as \texttt{WAIC} or \texttt{LOO} would penalize the model with additional intercept for cultural community for having additional parameters that do not help to improve predictive accuracy.

To get around this problem, we used a cross-validation procedure (see e.g., Stengelin et al., 2023). For each cultural setting, we randomly sampled a data set that was 5/6 the size of the full data set (training data). Then, we fit the model to this training data and used the estimated model parameters to predict the remaining 1/6 of the data (testing data). We then compared the model predictions from the different models by computing the mean difference between the true and predicted imprecision, over all trials in the testing data set. This approach gets around the problem mentioned above because the model predicts a new data set for which the individual random intercepts are unknown. Clustering by culture could therefore only be predicted by a model that included culture as a predictor. We repeated the cross-validation procedure 100 times and counted which model performed best most often.

@@
